### 2022-06-03 chapters/fairness/fif/introduction.tex
upto preliminaries in fairness

%\section{Introduction}
%\begin{itemize}
%	\item Bias -> fairness metric 
%	\item Fairness influence function (FIF)
%	\item Example
%	\item Contribution (1) framework (2) algorithm (3) experimental validation
%	\item Related works.
%	\item Notation paragraph
%\end{itemize}
The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.

>>>>

%\section{Introduction}
%\begin{itemize}
%	\item Bias -> fairness metric 
%	\item Fairness influence function (FIF)
%	\item Example
%	\item Contribution (1) framework (2) algorithm (3) experimental validation
%	\item Related works.
%	\item Notation paragraph
%\end{itemize}
\label{chapter:fif}
The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed machine learning classifiers often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.

### 2022-06-03 chapters/fairness/fvgm/introduction.tex
upto preliminaries in fairness

%\section{Introduction}
	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}. ML algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
%	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction. Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
	 The following example illustrates a plausible case of unfairness induced by ML algorithms. 
	\begin{example}\label{fvgm_example:intro}
		\normalfont
		Following Example~\ref{fairness_justicia_example:intro}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}). Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions. We generate 1000 samples from these distributions and use them to train a Support Vector Machine (SVM) classifier. The decision boundary of this classifier is $9.37I + 9.75F - 0.34A \ge 9.4$, where $A = 1$ denotes the sensitive group `age $ \ge $ 40'. This classifier selects an individual above and below $40$ years of age with probabilities $0.24$ and $0.86$, respectively. This illustrates a disparate treatment of individuals of two age groups by the SVM classifier. 
	\end{example}
%	The above example illustrates that even in simple scenarios, ML-based solutions can lead to disparate treatment. Motivated by studies indicating that such scenarios are rather not uncommon, there is a thrust in the research on the improvement of fairness in ML by proposing different notions of fairness definitions and fairness algorithms~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.

>>>>

%\section{Introduction}

\label{chapter:fvgm}
	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}. ML algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
%	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction. Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
	 The following example illustrates a plausible case of unfairness induced by ML algorithms. 
	\begin{example}\label{fvgm_example:intro}
		\normalfont
		Following Example~\ref{fairness_justicia_example:intro}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}). Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions. We generate 1000 samples from these distributions and use them to train a Support Vector Machine (SVM) classifier. The decision boundary of this classifier is $9.37I + 9.75F - 0.34A \ge 9.4$, where $A = 1$ denotes the sensitive group `age $ \ge $ 40'. This classifier selects an individual above and below $40$ years of age with probabilities $0.24$ and $0.86$, respectively. This illustrates a disparate treatment of individuals of two age groups by the SVM classifier. 
	\end{example}
%	The above example illustrates that even in simple scenarios, ML-based solutions can lead to disparate treatment. Motivated by studies indicating that such scenarios are rather not uncommon, there is a thrust in the research on the improvement of fairness in ML by proposing different notions of fairness definitions and fairness algorithms~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.

### 2022-06-03 chapters/fairness/justicia/introduction.tex
upto preliminaries in fairness

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
Machine learning is becoming the omnipresent technology of our time. Machine learning classifiers are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions. Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.

@@

\subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias. \red{In this chapter, we focus on a popular concept of fairness, known as group fairness.}\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}. Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}. Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.

>>>>

%\section{Introduction}
%\paragraph{Motivation: \red{ML} + Fairness}
\label{chapter:justicia}
Machine learning is becoming the omnipresent technology of our time. Machine learning classifiers are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions. Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.

@@

\subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias. In this chapter, we focus on a popular concept of fairness, known as group fairness.\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}. Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}. Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.

### 2022-06-03 chapters/fairness_background.tex
upto preliminaries in fairness

\part{Fairness in Machine Learning}
\chapter{Preliminaries}
In this chapter, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness. We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.

\subsection*{Dataset and Distribution}
We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$. Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $\{\mathbf{x}^{(i)}_1, \dots, \mathbf{x}^{(i)}_{\numnonsensitive}\} $. Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $\{\mathbf{a}^{(i)}_1, \dots, \mathbf{a}^{(i)}_{\numsensitive}\} $.  $y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.

We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.  Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. Thus, $ \mathcal{D} $ is the product distribution, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.

@@

We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $. Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $. Given this setup, we discuss different fairness metrics to compute bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.

\subsection*{Fairness Metrics}

A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}. In the following, we describe two well-known fairness definitions: group fairness and causal fairness. To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $. Alternatively, a fairness metric can be computed on a finite sampled dataset instead of the distribution. In that case, we use the notation $ f(\alg, \mathbf{D}) $, which is the bias of the classifier $ \alg $ on a dataset $ \mathbf{D} $. In the following, we discuss different fairness metrics.

@@

\paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.  Formally, the disparate impact of a classifier is 
\[
f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}.
\]

@@

\paragraph{Path-specific Causal Fairness ($ \mathsf{PCF} $).}
Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.   To this end, the path-specific causal fairness, abbreviated as $ \mathsf{PCF} $, of a classifier is \[
f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
\]

Therefore, path-specific cause fairness constrains that the prediction $ \hat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through mediator features $ \mediator $. Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.

@@

\subsection*{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form (CNF)} if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $. The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true). 
In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is computes the  probability of the satisfaction of the formula $\phi$ defined on \textit{quantified} Boolean variables. 
An SSAT formula is of the form
\begin{equation}\label{fairness_justicia_eq:ssat}
\Phi = Q_1\bool_1, \dots, Q_m \bool_m,\; \phi, 
\end{equation}
where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula. In case of a randomized quantifier $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $. In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.  Let $ \bool $ be the outermost variable in the prefix. The semantics of SSAT formulas are defined recursively in the following.
\begin{enumerate}
	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
\end{enumerate}
where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively. In this thesis, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT. In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.

@@

\subsection*{Stochastic Subset Sum Problem ({\stochastic})} 
Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ W_i \in \mathbb{Z} $ be the weight\change{Weight notation $ w $ to $ W $} of $ B_i $. Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$. Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}. The \textit{counting} version of the subset-sum problem counts all $ \mathbf{b} $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $. In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i \ge \tau $ where variables $ B_i $'s are stochastic. More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} W_iB_i \ge \tau] $.    Details of {\stochastic} are in Section~\ref{fvgm_sec:stochastic_sum_set_sum}.

\subsection*{Bayesian Network}
In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables. In this chapter, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them. 	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation. Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable. Edges $ \mathbf{E} \in \mathbf{V} \times \mathbf{V} $ imply conditional dependencies among variables. For each variable $ V_i \in \mathbf{V} $, let $ \parent(V_i) \subseteq \mathbf{V} \setminus \{V_i\} $ denote the set of parents of $ V_i $. Given $\parent(V_i)$ and parameters $\theta$, $ V_i $ is independent of its other non-descendant variables in $\graph$. Thus, for the assignment $ v_i $ of $ V_i $ and $ \mathbf{u} $ of $ \parent(V_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\mathbf{V}$ as:

@@

\subsection*{Global Sensitivity Analysis: Variance Decomposition}
Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
%\paragraph{Sensitivity Analysis:}
\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}. To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:

>>>>

\part{Fairness in Machine Learning}

In this part of the thesis, we discuss preliminaries on fairness in machine learning in Chapter~\ref{chapter_fairness_preliminaries}. Next, we discuss fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. Finally, we discuss identification of sources of unfairness in Chapter~\ref{chapter:fif}.

\chapter{Preliminaries}
\label{chapter_fairness_preliminaries}
In this chapter, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness. We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.

\subsection{Dataset and Distribution}
We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$. Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $. Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.  $y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.

We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.  Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. Thus, $ \mathcal{D} $ is the product distribution of all features, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.

@@

We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $. Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $. Given this setup, we discuss different fairness metrics to compute the bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.

\subsection{Fairness Metrics}

A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\widehat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}. In the following, we describe two well-known fairness definitions: group fairness and causal fairness. To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $. Alternatively, a fairness metric can be computed on a finite dataset instead of on the distribution. In that case, we use the notation $ f(\alg, \mathbf{D}) $, which is the bias of the classifier $ \alg $ on a dataset $ \mathbf{D} $.

@@

\paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact also belongs to independence measuring group fairness metrics. Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.  Formally, the disparate impact of a classifier is 
\[
f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\widehat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\widehat{Y} =1 | \mathbf{A} =  \mathbf{a}]}.
\]

@@

\paragraph{Path-specific Causal Fairness ($ \mathsf{PCF} $).}
Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 |\mathbf{A}=  \mathbf{a}] $. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.   To this end, the path-specific causal fairness, abbreviated as $ \mathsf{PCF} $, of a classifier is \[
f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
\]

Therefore, path-specific cause fairness constrains that the prediction $ \widehat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \widehat{Y} $ only through mediator features $ \mediator $. Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.

@@

\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form} (CNF) if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $. The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true). 
In contrast to the SAT problem, the \textit{stochastic Boolean satisfiability} (SSAT) problem~\cite{littman2001stochastic} computes the  probability of the satisfaction of the formula $\phi$ defined on an ordered set of \textit{quantified} Boolean variables. 
An SSAT formula is of the form
\begin{equation}\label{fairness_justicia_eq:ssat}
\Phi = Q_1\bool_1, \dots, Q_m \bool_m,\; \phi, 
\end{equation}
where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula. In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[B_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $. In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.  Let $ \bool $ be the outermost variable in the prefix. The semantics of SSAT formulas are defined recursively in the following.

\begin{enumerate}
	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
\end{enumerate}

where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively. In this thesis, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT. In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.

@@

\subsection{Stochastic Subset Sum Problem ({\stochastic})} 
Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ W_i \in \mathbb{Z} $ be the weight\change{Weight notation $ w $ to $ W $} of $ B_i $. Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$. Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}. The \textit{counting} version of the subset-sum problem counts all $ \mathbf{b} $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $. In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i \ge \tau $ where variables $ B_i $'s are stochastic. More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} W_iB_i \ge \tau] $.    Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.

\subsection{Bayesian Network}
In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables. In fairness verification with correlated features, we leverage an access to a Bayesian network on features $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them. A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation. Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable. Edges $ \mathbf{E} \in \mathbf{V} \times \mathbf{V} $ imply conditional dependencies among variables. For each variable $ V_i \in \mathbf{V} $, let $ \parent(V_i) \subseteq \mathbf{V} \setminus \{V_i\} $ denote the set of parents of $ V_i $. Given $\parent(V_i)$ and parameters $\theta$, $ V_i $ is independent of its other non-descendant variables in $\graph$. Thus, for the assignment $ v_i $ of $ V_i $ and $ \mathbf{u} $ of $ \parent(V_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\mathbf{V}$ as:

@@

\subsection{Global Sensitivity Analysis: Variance Decomposition}
Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
%\paragraph{Sensitivity Analysis:}
\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}. To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
